---
title: "03b"
author: "T. Griffin"
date: "2024-04-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Set Up
```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      fig.path = "./figures/03_phylogenetic-tree/")#send any figure output to this folder
```

# Set the seed
```{r set-the-seed}
#set seed
set.seed(238428)
```

# Goals:
1. load the fastree unrooted tree
2. Add tree to phyloseq object
3/ Visualize and inspect tree with ggtree
4. Prune asvs, if needed
5. Root our tree
6. Combine new tree with phyloseq object
7. Save two phyloseq object: 1. unrooted tree. 2. rooted tree

# Load packages
```{r load-packages}
pacman::p_load(tidyverse, ggtree, phyloseq, phytools, install = FALSE)

```

# Load data
```{r load-data}
#preprocessed phyloseq object
load("/local/workdir/tag92/git_repos/SalinityGradient_16S/data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

#load in the tree
unrooted_tree <- read.tree("/local/workdir/tag92/git_repos/SalinityGradient_16S/data/03_phylogenetic_tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)

```

# Merge Phyloseq
```{r}
#intuition check
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))

#Merge the tree with the phyloseq object
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)
unrooted_physeq
```

# Plot the tree using ggTree
```{r}
# Make a basic tree
kingdom_node_tree <- ggtree(unrooted_physeq) +
  #color tips by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) +
  scale_color_manual(values = c("goldenrod1", "pink", "red")) +
  #add title
  labs(title = "unrooted tree") +
  #add node label
  geom_text(aes(label=node), # horizontal adjustment 
            hjust = -0.5, vjust = -0.3, size = 2) +
  #move legend to the bottom
  theme(legend.position = "bottom")

```

# Evaluate Long branch
This part gets a bit manual
```{r eval-long-branch}
#inspect clades of the tree
#zoom in on origin tree: 2537

viewClade(kingdom_node_tree + labs(title = "unrooted tree node 5208") +
    geom_text(aes(label=ASV)),
          node = 5208)

unrooted_physeq %>%
  subset_taxa(ASV == "ASV_456") %>%
  tax_table() %>%
  data.frame()

```

# prune out the weird ASVs
```{r prune-ASV456}
#this will remove the strange ASVs
#create a new tree without it
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

unrooted_tree_removed456 <- unrooted_physeq %>%
  pop_taxa(., "ASV_456")

#intuiton check:this will give the number of taxa present after removal
ntaxa(unrooted_physeq) - ntaxa(unrooted_tree_removed456)

#visually inspect
ggtree(unrooted_tree_removed456) +
  geom_tippoint(mapping = aes(color = Kingdom))
```

#Midroot the tree
```{r midroot}
#midroot the tree: is the tree rooted?

new_unrooted_tree <- phy_tree(unrooted_tree_removed456)
is.rooted(new_unrooted_tree)
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)

# Is the new tree rooted?
is.rooted(midpoint_rooted_tree)

physeq_rm456 <- 
  raw_preprocessed_physeq %>%
  subset_taxa(ASV != "ASV_456")

# Merge tree with the new physeq_rm456
midroot_physeq_rm456 <- 
  merge_phyloseq(physeq_rm456, midpoint_rooted_tree)

midroot_physeq_rm456

ggtree(midroot_physeq_rm456) + 
  geom_tippoint(mapping = aes(color = Kingdom))
```

# Save to new phylo object
```{r save-physeq}
save(list = c("unrooted_tree_removed456", "midroot_physeq_rm456"),
     file = "/local/workdir/tag92/git_repos/SalinityGradient_16S/data/03_phylogenetic_tree/phytree_preprocessed_physeq.RData")

```

```{r session-info}
devtools::session_info()
```



