---
title: "Biodiversity analysis"
author: "T. Griffin"
date: "2024-03-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/04_Biodiversity/")
```

# Setting the Environment 

## Set the seed 
```{r set-seed}
set.seed(238428)
```

## Load Libraries 
```{r load-packages}
library(tidyverse)
library(devtools) 
library(patchwork)
#install.packages("iNEXT")
library(iNEXT)
library(phyloseq)

              
```

## Load in Data 
```{r load-data}
load("/local/workdir/tag92/git_repos/SalinityGradient_16S/data/02_PreProcessing/raw_physeq.RData")
raw_physeq

# Intiution Check 
min(sample_sums(raw_physeq))

metadata_df <-
  raw_physeq %>%
  sample_data() %>%
  data.frame()

head(metadata_df)

# Setting colors for stations 
station_colors <- c(
  "Shipping Channel" = "dodgerblue4",
  "Aransas Bay" = "dodgerblue2",
  "Copano West" = "#D9CC3C",
  "Copano East" = "#A0E0BA",
  "Mesquite Bay" = "#00ADA7")
```


# Goals

1. Calculate the Hill Diversity of the samples. 
2. Evaluate the rarefaction curves. 
3. Evaluate the Diversity values. 
4. Makes notes of specific samples and their seq depth. 

# Diversity Calculations with iNEXT 

```{r calc-div}
# prepare input data 
iNEXT_input_df <- 
  raw_physeq %>%
  otu_table() %>%
  data.frame()
# Quick check
dim(iNEXT_input_df)

# Run iNEXT: Calculate the Hill Numbers 
# Note that: Species in ROWS, Samples in COLUMNS 
# Remember to set the seed! 
#iNEXT_data <- iNEXT(iNEXT_input_df, q = c(0,1,2), datatype = "abundance")

# Save the file
#save(iNEXT_data, file = "/local/workdir/tag92/git_repos/SalinityGradient_16S/data/04_Biodiversity/iNEXT_data.RData")
```

# Evaluate the Diversity! 
```{r load-div}
load("/local/workdir/tag92/git_repos/SalinityGradient_16S/data/04_Biodiversity/iNEXT_data.RData")
str(iNEXT_data)
typeof(iNEXT_data)
```

# Plot Diversity 
```{r plot-rarefaction}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)

# END HERE 
```

```{r}
# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")
```

# Plot rarefaction with ggiNEXT
```{r ggiNEXT, fig.width=8, fig.height= 3.5}
#Plot rarefaction!
#rarefaction/extrapolation curve, type = 1

#order Q, when it is zero= it equals the richness/total number of taxa, when it equals 1, it equals the exponentialShannon/ number of common taxa, when it is 2, this is the inverse simpson/number of dominant taxa

ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  facet_wrap(~Order.q, scales = "fixed") + 
  scale_color_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors, guide = FALSE) + 
  scale_shape_manual(values = base::rep(17, nsamples(raw_physeq)),
                     guide = FALSE) + 
  theme(legend.position = "none")
```

# Manually Plot diversity
```{r rarefaction}
#qD is the actual diversity
#m = the different rarefactions (the hill number)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(station_colors = station_colors,
                          station = names(station_colors)),
            by = "station") 
# Inspect 
dim(iNEXT_manual_df)

# Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = station, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~station, scales = "free") + 
  scale_color_manual(values = station_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 
  
  # Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Extrapolation") %>%
  # Make the actual rarefaction plot with 
  # the # of sequences on the x-axis and diversity on the y-axis
  # You can choose to pick one diversity value or plot all three 
  ggplot(aes(x = m, y= qD, color = station, group = names)) + 
  # line 
  #geom_line() + 
  geom_point() + 
  # Challenge: Facet with the station
  facet_grid(Order.q~station, scales = "fixed") + 
  scale_color_manual(values = station_colors) + 
  theme(legend.position = "bottom")

#boxplots
station_names <- c("Copano\nWest", "Copano\nEast","Mesquite\nBay",
                   "Aransas\nBay", "Shipping\nChannel")

names(station_names) <- c("Copano West", "Copano East","Mesquite Bay",
                          "Aransas Bay", "Shipping Channel")

# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(station_names = station_names, 
                       station = names(station_names)), 
                       by = "station")

# Check it 
head(obs_div_df)

obs_div_df %>%
  ggplot(aes(x = station_names, y = qD, fill = station, color = station)) + 
  facet_wrap(~Order.q, scales = "free") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

#Diversity vs salinity

```{r environmental-pairs, fig.width =8, }
head(metadata_df)
#pull out environmental variables
env_df <- 
  metadata_df %>%
  dplyr::select(names, water_tempC:DO_mgL)
#inspect
head(env_df)

  #in dyplr select gives you columns
#plot the correlations
pairs(dplyr::select(env_df, -names), upper.panel = NULL)
```


Diversity vs Salinity
```{r div-vs-salinity, fig.height=6, fig.width =6}
#PSU = practical salinity unit
#1 PSU = 1 g of salt per 1000 grams of water
#1 PSU = 1 ppt (part per thousand)

#freshwater = 0-0;5
#brackish = 0.5 - 30
#Marine = >30

view(iNEXT_manual_df)

iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  ggplot(aes(x = salinity_psu, y = qD)) +
  facet_grid(~Order.q, scales = "free") +
  geom_point(aes(color = station)) +
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) +
  labs(x = "Salinity (PSU)", y = "Num. of ASVs") +
  scale_color_manual(values = station_colors) +
theme(legend.position = "bottom")
```


```{r Session Information}
devtools::session_info()
